\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2010-08-09 papers/jss/V2Revision/dpx4Rev.Rnw
Cleaned up papers directory
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@437 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\title{\pkg{depmixS4} : An \proglang{R}-package for Hidden Markov Models}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Ingmar Visser, Maarten Speekenbrink} %% comma-separated
\Plaintitle{depmixS4: An R-package for Hidden Markov Models} %% without formatting

@@

	\pkg{depmixS4} implements a general framework for defining and
	estimating dependent mixture models in the \proglang{R}
	programming language \citep{R2010}.  This includes standard Markov
	models, latent/hidden Markov models, and latent class and finite
	mixture distribution models.  The models can be fitted on mixed
	multivariate data with distributions from the \code{glm} family,
	the (logistic) multinomial, or the multivariate normal
	distribution.  Other distributions can be added easily, and an
	example is provided with the {\em exgaus} distribution.
	Parameters are estimated by the EM algorithm or, when (linear)
	constraints are imposed on the parameters, by direct numerical
	optimization with the \pkg{Rsolnp} or \pkg{Rdonlp2} routines.

}

@@

%% \usepackage{Sweave}

@@

%set width of figures produced by Sweave
\setkeys{Gin}{width=0.8\textwidth}

@@

<<echo=FALSE,results=hide>>=	
options(prompt="R> ", continue="+	",width=70, useFancyQuotes=FALSE,
digits=4)
library(depmixS4)
@

@@

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes; see \citet{Wickens1982},
for an overview, and e.g. \citet{Schmittmann2006}, for a recent
application.  In economics, latent Markov models are so-called regime
switching models (see e.g., \citealp{Kim1994} and
\citealp{Ghysels1994}).  Further applications include speech
recognition \citep{Rabiner1989}, EEG analysis \citep{Rainer2000}, and
genetics \citep{Krogh1998}.  In these latter areas of application,
latent Markov models are usually referred to as hidden Markov models.
See for example \citet{Fruhwirth2006} for an overview of hidden Markov
models with extensions.  Further examples of applications can be found
in e.g.\ \citet[][chapter~1]{Cappe2005}.  A more gentle introduction
into hidden Markov models with applications is the book by
\citet{Zucchini2009}.

@@

\begin{figure}[!Htbp]
\begin{center}
<<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=	
data(speed)
plot(as.ts(speed[1:168,]),main="Speed-accuracy trade-off")
@
	\caption{Response times (rt), accuracy (corr) and pay-off 
	values (Pacc) for the first series of responses in dataset 
	\code{speed}.}
	\label{fig:speed}
\end{center}
\end{figure}

@@

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM
algorithm, parameters are estimated by iteratively maximizing the
expected joint log-likelihood of the parameters given the observations and
states.  Let $\greekv{\theta} = (\greekv{\theta}_1,
\greekv{\theta}_2,\greekv{\theta}_3)$ be the general parameter vector
consisting of three subvectors with parameters for the prior model,
transition model, and response models respectively.  The joint
log-likelihood can be written as:
\begin{equation}
\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(\vc{O}_t|S_t,\vc{z}_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
Expectation step, we replace these with their expected values given a set of 
(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $\vc{O}_{1:T}$. 
The expected log-likelihood:
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta})),
\end{equation}
can be written as:
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
\log \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
\end{multline}
where the expected values $\xi_t(j,k) = P(S_t = k, S_{t-1} =
j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) =
P(S_t = j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ can be
computed effectively by the forward-backward algorithm \citep[see
e.g.,][]{Rabiner1989}.  The Maximization step consists of the
maximization of (\ref{eq:Q}) for $\greekv{\theta}$.  As the right hand
side of (\ref{eq:Q}) consists of three separate parts, we can maximize
separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and
$\greekv{\theta}_3$.  In common models, maximization for
$\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the
\code{nnet.default} routine in the \pkg{nnet} package
\citep{Venables2002}, and maximization for $\greekv{\theta}_3$ by the
standard \code{glm} routine.  Note that for the latter maximization,
the expected values $\gamma_t(j)$ are used as prior weights of the
observations $O^k_t$.

@@

Two steps are involved in using \pkg{depmixS4} which are illustrated
below with examples:
\begin{enumerate}

	\item model specification with function \code{depmix} (or with
	\code{mix} for latent class and finite mixture models, see example
	below on adding covariates to prior probabilities)

@@

These statistics can also be extracted using \code{logLik}, \code{AIC}
and \code{BIC}, respectively.  By comparison, a 1-state model for
these data, i.e. assuming there is no mixture, has a log-likelihood of
$-305.33$, and 614.66, and 622.83 for the AIC and BIC respectively.
Hence, the 2-state model fits the data much better than the 1-state
model.  Note that the 1-state model can be specified using \code{mod
<- depmix(rt~1, data=speed, nstates=1)}, although this model is
trivial as it will simply return the mean and standard deviation of
the \code{rt} variable.

@@

Multivariate data can be modelled by providing a list of formulae as
well as a list of family objects for the distributions of the various
responses.  In above examples we have only used the response times
which were modelled as a Gaussian distribution.  The accuracy
variable in the \code{speed} data can be modelled with a multinomial
by specifying the following:
<<>>=
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
    family=list(gaussian(), multinomial("identity")),
		transition=~scale(Pacc),instart=runif(2))
fm <- fit(mod,verbose=FALSE)
@

@@

To see which parameters are fixed (by default only baseline parameters
in the multinomial logistic models for the transition models and the
initial state probabilities model):
<<term=FALSE>>=
setpars(mod, getpars(mod,which="fixed"))
@
When fitting constraints it is useful to have good starting values 
for the parameters and hence we first fit the following model without
constraints:
<<>>=
trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
	  nstates=2,family=list(gaussian(),multinomial("identity")),
	  trstart=trst,instart=c(0.99,0.01))
fm1 <- fit(mod,verbose=FALSE)
@

After this, we use the fitted values from this model to constrain the
regression coefficients on the transition matrix (parameters number~6
and~10):
<<term=FALSE>>=
pars <- c(unlist(getpars(fm1)))
pars[6] <- pars[10] <- 11
pars[1]=0
pars[2]=1
pars[13] <- pars[14] <- 0.5
fm1 <- setpars(mod,pars)
# start with fixed and free parameters
conpat <- c(0,0,rep(c(0,1),4),1,1,0,0,1,1,1,1)
# constrain the beta's on the transition parameters to be equal
conpat[6] <- conpat[10] <- 2
@
<<echo=TRUE,results=hide>>=
fm2 <- fit(fm1,equal=conpat)
@

@@

The balance scale task is a famous task for testing cognitive
strategies developed by Jean Piaget \citep[see][]{Siegler1981}.
Figure~\ref{fig:balance} provides an example of a balance scale item.
Participants' task is to say to which side the balance will tip when
released, or alternatively, whether it will stay in balance.  The item
shown in Figure~\ref{fig:balance} is a so-called distance item: the
number of weights placed on each side is equal, and only the distance
of the weights to the fulcrum differs between each side.

@@

User-defined distributions should extend the \code{response}-class and
have the following slots:
\begin{enumerate}
	\item \code{y}: the response variable
	\item \code{x}: the design matrix, possibly only an intercept
	\item \code{parameters}: a named list with the coefficients and possibly 
	other parameters (e.g., the standard deviation in the normal 
	response model)
	\item \code{fixed}: a vector of logicals indicating whether parameters are 
	fixed
	\item \code{npar}: numerical indicating the number of parameters of the model
\end{enumerate}

@@

The so-defined class now needs a number of methods: 
\begin{enumerate}
	\item constructor: function to create instances of the class 
	with starting values
	\item \code{show}: to print the model to the terminal
	\item \code{dens}: the function that computes the density of the responses
	\item \code{getpars} and \code{setpars}: to get and set parameters 
	\item \code{predict}: to generate predict'ed values 
	\item \code{fit}: function to fit the model using posterior weights (used 
	by the EM algorithm)
\end{enumerate}

Only the constructor and the \code{fit} methods are provided here; the
complete code can be found in the help file of the \code{makeDepmix}
function.  The example with the exgaus distribution uses the
\pkg{gamlss} and \pkg{gamlss.distr} packages \citep{Rigby2005,
Stasinopoulos2007,Stasinopoulos2009a,Stasinopoulos2009b} for computing
the \code{dens}ity and for \code{fit}ting the parameters.

The constructor method return an object of class \code{exgaus}, and is
defined as follows:
<<results=hide>>=
require(gamlss)
require(gamlss.dist)
@
<<term=FALSE>>=
setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
  standardGeneric("exgaus"))

setMethod("exgaus", 
  signature(y="ANY"), 
  function(y, pstart=NULL, fixed=NULL, ...) {
    y <- matrix(y,length(y))
    x <- matrix(1) 
    parameters <- list()
    npar <- 3
    if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
    if(!is.null(pstart)) {
      if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
      parameters$mu <- pstart[1]
      parameters$sigma <- log(pstart[2])
      parameters$nu <- log(pstart[3])
    }
    mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
    mod
  }
)
@

@@

setMethod("predict","exgaus", 
    function(object) {
        ret <- object@parameters$mu
        return(ret)
    }
)
@

The \code{fit} method is defined as follows: 
<<echo=FALSE>>=
options(width=60)
@
<<>>=
setMethod("fit","exgaus",
  function(object, w) {
    if(missing(w)) w <- NULL
    y <- object@y

	fit <- gamlss(y~1,weights=w,family=exGAUS(),
      control=gamlss.control(n.cyc=100,trace=FALSE),
      mu.start=object@parameters$mu, 
	  sigma.start=exp(object@parameters$sigma), 
      nu.start=exp(object@parameters$nu))

	pars <- c(fit$mu.coefficients, fit$sigma.coefficients, 
		fit$nu.coefficients)
    object <- setpars(object,pars)
    object
  }
)
@

@@

We first create all the response models that we need as a double list: 
<<echo=FALSE>>=
options(width=80)
@
<<>>=
rModels <- list()
rModels[[1]] <- list()
rModels[[1]][[1]] <- exgaus(speed$rt, pstart=c(5,.1,.1))
rModels[[1]][[2]] <- GLMresponse(formula=corr~1, data=speed, 
	family=multinomial(), pstart=c(0.5,0.5))

rModels[[2]] <- list()
rModels[[2]][[1]] <- exgaus(speed$rt, pstart=c(6,.1,.1))
rModels[[2]][[2]] <- GLMresponse(formula=corr~1, data=speed, 
	family=multinomial(), pstart=c(.1,.9))
@

Next, we define the transition and prior probability models using the 
\code{transInit} function (which produces a transInit model, which also extends 
the response class): 
<<echo=FALSE>>=
options(width=70)
@
<<>>=
trstart=c(0.9,0.1,0.1,0.9)
transition <- list()
transition[[1]] <- transInit(~Pacc,nst=2, data=speed, 
	pstart=c(0.9,0.1,0,0))
transition[[2]] <- transInit(~Pacc,nst=2, data=speed, 
	pstart=c(0.1,0.9,0,0))
inMod <- transInit(~1, ns=2, pstart=c(0.1,0.9), data=data.frame(1))
@

Finally, we put everything together using \code{makeDepmix} and fit 
the model: 
<<>>=
mod <- makeDepmix(response=rModels, transition=transition,
    prior=inMod, stat=FALSE)
fm <- fit(mod, verbose=FALSE)
@

@@

\pkg{depmixS4} provides a flexible framework for fitting dependent
mixture models for a large variety of response distributions.  It can
also fit latent class regression and finite mixture models, although
it should be noted that more specialized packages are available for
this such as \pkg{FlexMix} \citep{Leisch2004}.  The package is
intended for modelling (individual) time series data with the aim of
characterizing the transition processes underlying the data.  The
possibility to use covariates on the transition matrix greatly
enhances the flexibility of the model.  The EM algorithm uses a very
general interface that allows easy addition of new response models.

We are currently working on implementing the gradients for response
and transition models with two goals in mind.  First, to speed up
(constrained) parameter optimization using \pkg{Rdonlp2} or
\pkg{Rsolnp}.  Second, analytic gradients are useful in computing the
Hessian of the estimated parameters so as to arrive at standard errors
for those.  We are also planning to implement goodness-of-fit
statistics \citep{Titman2008}.

@@

Ingmar Visser was supported by an EC Framework 6 grant, project 516542
(NEST).  Maarten Speekenbrink was supported by ESRC grant
RES-062-23-1511 and the ESRC Centre for Economic Learning and Social
Evolution (ELSE).  Han van der Maas provided the speed-accuracy data
\citep{Dutilh2010} and thereby necessitated implementing models with
time-dependent covariates.  Brenda Jansen provided the balance scale
data set \citep{Jansen2002} which was the perfect opportunity to test
the covariates on the prior model parameters.  The examples in the
help files use both of these data sets.

>>>>

\title{\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Ingmar Visser, Maarten Speekenbrink} %% comma-separated
\Plaintitle{depmixS4: An R-package for hidden Markov models} %% without formatting

@@

	\pkg{depmixS4} implements a general framework for defining and
	estimating dependent mixture models in the \proglang{R}
	programming language \citep{R2009}.  This includes standard Markov
	models, latent/hidden Markov models, and latent class and finite
	mixture distribution models.  The models can be fitted on mixed
	multivariate data with distributions from the \code{glm} family,
	the (logistic) multinomial, or the multivariate normal distribution.
	Other distributions can be added easily, and an example is
	provided with the exgaus distribution.  Parameters are estimated by
	the EM algorithm or, when (linear) constraints are imposed on the
	parameters, by direct numerical optimization with the
	\pkg{Rsolnp} or \pkg{Rdonlp2} routines.}

@@

%\usepackage{Sweave}

@@

%set width of figures produced by Sweave
\setkeys{Gin}{width=0.9\textwidth}

@@

@@

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes (see \citealp{Wickens1982},
for an overview, and e.g. \citealp{Schmittmann2006}, for a recent
application).  In economics, latent Markov models are so-called regime
switching models (see e.g., \citealp{Kim1994} and
\citealp{Ghysels1994}).  Further applications include speech
recognition \citep{Rabiner1989}, EEG analysis \citep{Rainer2000}, and
genetics \citep{Krogh1998}.  In these latter areas of application,
latent Markov models are usually referred to as hidden Markov models.
See for example \citet{Fruhwirth2006} for an overview of hidden Markov
models with extensions.  Further examples of applications can be found
in e.g.\ \citet[][chapter~1]{Cappe2005}.

@@

\begin{figure}[htbp]
\begin{center}
<<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=	
.Options$digits<-3
library(depmixS4)
data(speed)
plot(as.ts(speed[1:168,]),main="Speed-accuracy trade-off")
@
	\caption{Response times (rt), accuracy (corr) and pay-off 
	values (Pacc) for the first series of responses in dataset 
	\code{speed}.}
	\label{fig:speed}
\end{center}
\end{figure}

@@

%transition probability functions $a_{ij}$ and the initial state
%probability functions $\greekv{\pi}$ may depend on covariates as well
%as the response distributions $\vc{b_{j}^{k}}$.  See for example
%\citet{Fruhwirth2006} for an overview of hidden Markov models with
%extensions.

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM
algorithm, parameters are estimated by iteratively maximising the
expected joint log-likelihood of the parameters given the observations and
states.  Let $\greekv{\theta} = (\greekv{\theta}_1,
\greekv{\theta}_2,\greekv{\theta}_3)$ be the general parameter vector
consisting of three subvectors with parameters for the prior model,
transition model, and response models respectively.  The joint
log-likelihood can be written as:
\begin{equation}
\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(\vc{O}_t|S_t,\vc{z}_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
Expectation step, we replace these with their expected values given a set of 
(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $\vc{O}_{1:T}$. 
The expected log-likelihood:
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta})),
\end{equation}
can be written as:
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
\log \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
\end{multline}
where the expected values $\xi_t(j,k) = P(S_t = k, S_{t-1} =
j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) =
P(S_t = j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ can be
computed effectively by the forward-backward algorithm \citep[see
e.g.,][]{Rabiner1989}.  The Maximisation step consists of the
maximisation of (\ref{eq:Q}) for $\greekv{\theta}$.  As the right hand
side of (\ref{eq:Q}) consists of three separate parts, we can maximise
separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and
$\greekv{\theta}_3$.  In common models, maximisation for
$\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the
\code{nnet.default} routine in the \pkg{nnet} package
\citep{Venables2002}, and maximisation for $\greekv{\theta}_3$ by the
standard \code{glm} routine.  Note that for the latter maximisation,
the expected values $\gamma_t(j)$ are used as prior weights of the
observations $O^k_t$.

@@

Two steps are involved in using \pkg{depmixS4} which are illustrated
below with examples:
\begin{enumerate}
	\item model specification with function \code{depmix} (or with \code{mix}
	for latent class and finite mixture models, see example below on adding
	covariates to prior probabilities)

@@

These statistics can also be extracted using \code{logLik}, \code{AIC}
and \code{BIC}, respectively.  By comparison, a 1-state model for
these data, i.e. assuming there is no mixture, has a log-likelihood of
$-305.33$, and 614.66, and 622.83 for the AIC and BIC respectively.
Hence, the 2-state model fits the data much better than the 1-state
model.  Note that the 1-state model can be specified using \code{mod <-
depmix(rt~1, data=speed, nstates=1)}, although this model is trivial
as it will simply return the mean and the sd of the rt variable.

@@

Multivariate data can be modelled by providing a list of formulae as
well as a list of family objects for the distributions of the various
responses.  In above examples we have only used the response times
which were modelled as a Gaussian distribution.  The accuracy
variable in the \code{speed} data can be modelled with a multinomial
by specifying the following:
<<>>=
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
    family=list(gaussian(),multinomial("identity")),
    transition=~scale(Pacc),instart=runif(2))
fm <- fit(mod,verbose=FALSE)
@

@@

To see which parameters are fixed (by default only baseline parameters
in the multinomial logistic models for the transition models and the
initial state probabilities model):
<<term=FALSE>>=
setpars(mod, getpars(mod,which="fixed"))
@
When fitting constraints it is useful to have good starting values 
for the parameters and hence we first fit the following model without
constraints:
<<>>=
trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
	  nstates=2,family=list(gaussian(),multinomial("identity")),
	  trstart=trst,inst=c(0.99,0.01))
fm1 <- fit(mod,verbose=FALSE)
@

After this, we use the fitted values from this model to constrain the
regression coefficients on the transition matrix (parameters number~6
and~10):
<<term=FALSE>>=
pars <- c(unlist(getpars(fm1)))
pars[6] <- pars[10] <- 11
pars[1]=0
pars[2]=1
pars[13] <- pars[14] <- 0.5
fm1 <- setpars(mod,pars)
# start with fixed and free parameters
conpat <- c(0,0,rep(c(0,1),4),1,1,0,0,1,1,1,1)
# constrain the beta's on the transition parameters to be equal
conpat[6] <- conpat[10] <- 2
fm2 <- fit(fm1,equal=conpat)
@

@@

The balance scale taks is a famous task for testing cognitive
strategies developed by Jean Piaget \citep[see][]{Siegler1981}.
Figure~\ref{fig:balance} provides an example of a balance scale item.
Participants' task is to say to which side the balance will tip when
released, or alternatively, whether it will stay in balance.  The item
shown in Figure~\ref{fig:balance} is a so-callled distance item: the
number of weights placed on each side is equal, and only the distance
of the weights to the fulcrum differs between each side.

@@

User-defined distributions should extend the \code{response}-class and
have the following slots:
\begin{enumerate}
	\item y: the response variable
	\item x: the design matrix, possibly only an intercept
	\item paramaters: a named list with the coefficients and possibly 
	other parameters (e.g., the standard deviation in the Gaussian 
	response model)
	\item fixed: a vector of logicals indicating whether parameters are 
	fixed
	\item npar: numerical indicating the number of parameters of the model
\end{enumerate}

@@

The so-defined class now needs a number of methods: 
\begin{enumerate}
	\item constructor: function to create instances of the class 
	with starting values
	\item show method: to print the model to the terminal
	\item dens: the function that computes the density of the responses
	\item getpars and setpars: to get and set parameters 
	\item predict: to generate predict'ed values 
	\item fit: function to fit the model using posterior weights (used 
	by the EM algorithm)
\end{enumerate}

Only the constructor and the fit methods are provided here; the
complete code can be found in the help file of the \code{makeDepmix}
function.  The \code{exgaus} example uses the \pkg{gamlss} and
\pkg{gamlss.distr} packages
\citep{Stasinopoulos2009a,Stasinopoulos2009b} for computing the
\code{dens}ity and for \code{fit}ting the parameters.

The constructor method return an object of class \code{exgaus}, and is
defined as follows:
<<term=FALSE>>=
setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
  standardGeneric("exgaus"))

require(gamlss)
require(gamlss.dist)

setMethod("exgaus", 
  signature(y="ANY"), 
  function(y,pstart=NULL,fixed=NULL, ...) {
    y <- matrix(y,length(y))
    x <- matrix(1) 
    parameters <- list()
    npar <- 3
    if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
    if(!is.null(pstart)) {
      if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
      parameters$mu <- pstart[1]
      parameters$sigma <- log(pstart[2])
      parameters$nu <- log(pstart[3])
    }
    mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
    mod
  }
)

@

@@

setMethod("predict","exgaus", 
    function(object) {
        ret <- object@parameters$mu
        return(ret)
    }
)

@

The \code{fit} method is defined as follows: 
<<>>=
setMethod("fit","exgaus",
  function(object,w) {
    if(missing(w)) w <- NULL
    y <- object@y
    fit <- gamlss(y~1,weights=w,family=exGAUS(),
      control=gamlss.control(n.cyc=100,trace=FALSE),
      mu.start=object@parameters$mu,
      sigma.start=exp(object@parameters$sigma),
      nu.start=exp(object@parameters$nu))
    pars <- c(fit$mu.coefficients,fit$sigma.coefficients,fit$nu.coefficients)
    object <- setpars(object,pars)
    object
  }
)
@

@@

We first create all the response models that we need as a double list: 
<<>>=
rModels <- list( list( 
		exgaus(speed$rt,pstart=c(5,.1,.1)),	
		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))),
	list(
		exgaus(speed$rt,pstart=c(6,.1,.1)),
		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))))
@

Next, we define the transition and prior probability models using the 
\code{transInit} function (which produces a transInit model, which also extends 
the response class): 
<<>>=
trstart=c(0.9,0.1,0.1,0.9)
transition <- list()
transition[[1]] <- transInit(~Pacc,nst=2,data=speed,pstart=c(0.9,0.1,0,0))
transition[[2]] <- transInit(~Pacc,nst=2,data=speed,pstart=c(0.1,0.9,0,0))
inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
@

Finally, we put everything together using \code{makeDepmix} and fit 
the model: 
<<>>=
mod <- makeDepmix(response=rModels,transition=transition,
    prior=inMod,stat=FALSE)
fm <- fit(mod,verbose=FALSE)
@

@@

\pkg{depmixS4} provides a flexible framework for fitting dependent
mixture models for a large variety of response distributions.  It can
also fit latent class regression and finite mixture models, although
it should be noted that more specialized packages are available for
this such as \pkg{FlexMix} \citep{Leisch2004}.  The package is
intended for modelling of (individual) time series data with the aim of
characterizing the transition processes underlying the data.  The
possibility to use covariates on the transition matrix greatly
enhances the flexibility of the model.  The EM algorithm uses a very
general interface that allows easy addition of new response models.

We are currently working on implementing the gradients for response
and transition models with two goals in mind.  First, to speed up
(constrained) parameter optimization using \pkg{Rdonlp2}.  Second,
analytic gradients are useful in computing the Hessian of the
estimated parameters so as to arrive at standard errors for those.  We
are also planning to implement goodness-of-fit statistics
\citep{Titman2008}.

@@

Ingmar Visser was supported by an EC Framework 6 grant, project 516542
(NEST).  Maarten Speekenbrink was supported by ESRC grant RES-062-23-1511 
and the ESRC Centre for Economic Learning and Social Evolution (ELSE).  
Han van der Maas provided the speed-accuracy data \cite{Dutilh2010} and thereby
neccessitated implementing models with time-dependent covariates.
Brenda Jansen provided the balance scale data set \citep{Jansen2002}
which was the perfect opportunity to test the covariates on the prior
model parameters.  The examples in the help files use both of these
data sets.
